Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/appendix.tex
work on proof of adequacy and move it to concrete model; work on dyn in concrete model

  We define $\push_{Fc} : \mathbb{N} \times P_{A} \to \mathbb{N} \times P_{A'}$.
  Since the product of commutative monoids coincides with the coproduct, it
  suffices by the universal property of the coproduct to consider the two cases separately.

  Given $n \in \mathbb{N}$ we define the push to be $n$.
  Then the required square is $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ which exists by
  the definition of a step-1 model.
  On the other hand, given $a \in P_{A}$, we define the push to be $\push_c(a)$
  and then the required exists by the push property for $c$.
  Defining $\pull_{Fc}$ is done similarly.

@@

  % Likewise, given push-pull structures 
  % $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
  % $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
  % we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
  % with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
\end{lemma}
\begin{proof}
We define the push function for $\piv_{c_1 \times c_2}$ by 
$\push(\delta_1, \delta_2) = (\push_{c_1}(\delta_1), \push_{c_2}(\delta_2))$ and likewise for $\pull$.
The push-pull property holds because it holds for each component.
\end{proof}

@@

      \[ (A_1, P_{A_1}) \times (A_2, P_{A_2}) =
        (A_1 \times A_2, P_{A_1} \times P_{A_2}). \]

      where $\oplus$ is the coproduct in the category of monoids.

@@

      \[ (A, P_A, \ptb_A) \arr (B, P_B, \ptb_B) = 
        (A \arr B, P_A \times P_B, \ptb_{A \arr B}), \]

@@

    \eric{Make sure this part is correct}
    We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
    to a perturbation in $P_{U(A \to B)}$, as follows:
    Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
    $(n, \id_{P_A}, b) \in P_{U(A \to B)}$.
    We need to show that
%
    \[ \ptb_{U(A \arr B)}(\id \tok (n, b)) = \id \tok \ptb_{UB}(n, b). \]
%
    Since the product and coproduct coincide in the category of commutative monoids,
    and since $\mathbb{N}$ is generated by the element $1$,
    it suffices by the universal property of the coproduct to show
%
    \[ \ptb_{U(A \arr B)}(\id \tok (1, \id_{P_B})) = \id \tok \ptb_{UB}(1, \id_{P_B}), \]
%
    and
%
    \[ \ptb_{U(A \arr B)}(\id \tok (0, b)) = \id \tok \ptb_{UB}(0, b). \]

    For the former, the LHS is
%
    \begin{align*}
      \ptb_{U(A \arr B)}&(\id \tok (1, \id_{P_B})) \\ 
      &= \ptb_{U(A \arr B)}(1, \id_{P_A}, \id_{P_B}) \\ 
      &= \delta_{U(A \arr B)}^* \circ U(\ptb_A(\id_{P_A}) \arr \ptb_B(\id_{P_B})) \\
      %&= \delta_{U(A \arr B)}^* \circ U(\id_A \arr \id_B) \\
      &= \delta_{U(A \arr B)}^*
    \end{align*}
%
    and we need to show that this is equal to 
    $\id \tok \ptb_{UB}(1, \id_{P_B}) = \id \tok \delta_{UB}^*$.
    But this holds because $\delta_{UB}^*$ commutes with all computation morphisms,
    in particular, with bind and application.

    For the latter, we need to show
%
    \[ U(\id \arr \ptb_B(b)) = \id \tok U(\ptb_B(b)), \]
%
    which holds by definition of $\tok$.

    Given a perturbation $q = (n, a) \in P_{FA}$ we define
    $(n, a) \tok \id = (n, a, \id_{P_B}) \in P_{U(A \to B)}$.
    We need to show that
  %
    \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
  %
    By similar reasoning to the above, it will suffice to show that

    \[ \delta_{U(A \arr B)}^* = \delta_{FA}^* \tok \id \]

    And this follows from the fact that $\delta_{FA}^*$ commutes with all computation morphisms.

    We can similarly construct the Kleisli product operation on perturbations.
    Given a perturbation $(n, a_1, a_2) \in P_{F(A_1 \times A_2)}$ we define
    $(n, a_1, a_2) \timesk \id = (n, a_1, \id_{P_{A_2}})$,
    and likewise we define $\id \timesk (n, a_1, a_2) = (n, \id_{P_{A_1}}, a_2)$.

@@

\begin{definition}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:

  \begin{center}
    \begin{tabular}{ m{9em} m{9em} } 
      \begin{tikzcd}[ampersand replacement=\&]
        A \& {A'} \\
        A \& {A'}
        \arrow["\delta^l_1"', from=1-1, to=2-1]
        \arrow["\delta^r_1", from=1-2, to=2-2]
        \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
      &
      \begin{tikzcd}[ampersand replacement=\&]
        A \& {A'} \\
        A \& {A'}
        \arrow["\delta^l_2"', from=1-1, to=2-1]
        \arrow["\delta^r_2", from=1-2, to=2-2]
        \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
    \end{tabular}
  \end{center}

  We make the analogous definition for computation relations $d, d' : B \rel B'$.
\end{definition}

\begin{lemma}\label{lem:left-rep-by-same-morphism}
  Let $c, c' : A \rel A'$. If $c$ and $c'$ are both quasi-left-representable
  by the same $f$, then $c \qordeq c'$.
\end{lemma}
\begin{proof}
  By $\upr$ for $c'$, there exists a perturbation $\delle_{c'}$ and a square
  $\upr_{c'} : \delle_{c'} \ltdyn_{c'}^{r(A)} f$.

@@

%%%%%%%%%%%%%%%
% Composition %
%%%%%%%%%%%%%%%
\begin{lemma}\label{lem:representation-comp}
  In the below, let $c : A \rel A'$ and $c' : A' \rel A''$ and $d : B \rel B'$ and $d' : B' \rel B''$.
  \begin{enumerate}
    \item Given left-representation structures $\rho^L_c$ for $c$ and $\rho^L_{c'}$ for $c'$,
    we can define a left-representation structure for the composition $c \comp c'$
    \item Given right-representation structures $\rho^R_d$ for $d$ and $\rho^R_{d'}$ for $d'$,
    we can define a right-representation structure for the composition $d \comp d'$
    \item Given right-representation structures $\rho^R_{Fc}$ for $Fc$ and $\rho^R_{Fc'}$ for $Fc'$,
    we can define a right-representation structure $\rho^R_{F(c \comp c')}$ for $F(c \comp c')$.
    \item Given left-representation structures $\rho^L_{Ud}$ for $Ud$ and $\rho^L_{Ud'}$ for $Ud'$,
    we can define a left-representation structure $\rho^L_{U(d \comp d')}$ for $U(d \comp d')$.
  \end{enumerate}
  \eric{Do we need parts 3 and 4?}

@@

    First, we claim that $F(c \comp c')$ and $Fc \comp Fc'$ are
    both left-represented by $Fe_{c'} \circ Fe_c$.
    Indeed, we have by part (1) that $e_c' \circ e_c$ left-represents $c \comp c'$,
    and then since $F$ preserves perturbations and squares it follows that
    $F(e_c' \circ e_c) = Fe_{c'} \circ Fe_c$ quasi-left-represents $F(c \comp c')$.
    On the other hand, we also know that $Fe_c$ quasi-left-represents $Fc$ and $Fe_{c'}$
    left-represents $Fc'$, so their composition quasi-left-represents $Fc \comp Fc'$.
    Thus, there is a square $\alpha$ of the form

@@

  % TODO check this
  We define $\rho^R_{F(c_1 \times c_2)}$ as follows:
  \begin{itemize}
    \item $p_{F(c_1 \times c_2)} = (p_{Fc_1} \timesk A_2) \circ (A_1' \timesk p_{Fc_2})$
    \item $\dellp_{F(c_1 \times c_2)} = (\dellp_{Fc_1} \timesk A_2) \circ (A_1 \timesk \dellp_{Fc_2})$
    \item $\delrp_{F(c_1 \times c_2)} = (\delrp_{Fc_1} \timesk A_2') \circ (A_1' \timesk \delrp_{Fc_2})$
    \item The commuting squares are obtained via the functorial action of $\timesk$ on the squares for $Fc_1$ and $Fc_2$.
    % TODO elaborate?
  \end{itemize}
\end{proof}

@@

\subsection{Defining an Extensional Model}\label{sec:extensional-construction-appendix}

@@

  Recall the extensional model defined in Section \ref{sec:extensional-model-definition}.

  We first establish the representability properties for this model.
  We show the left-representability squares; the right-representability squares are dual.

@@

@@

\section{Adequacy}\label{sec:appendix-adequacy}

In this section, we show an adequacy result for the extensional model of GTT we obtained by
applying the abstract construction introduced in Section
\ref{sec:extensional-model-construction} to the concrete model

First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$.
We write that $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.

Recall that $\ltls$ denotes the relation on value morphisms defined as the bisimilarity-closure
of the intensional error-ordering on morphisms.
That is, we have $f \ltls g$ iff there exists $f'$ and $g'$ with

\[ f \bisim f' \le g' \bisim g. \]

The result we would like to show is as follows:
\begin{lemma}
If $f \ltls g : \li \mathbb{N}$, then:
\begin{itemize}
  \item If $f \da n$ then $g \da n$.
  \item If $g \da \mho$ then $f \da \mho$.
  \item If $g \da n$ then $f \da n$.
\end{itemize}
\end{lemma}

Unfortunately, this is actually not provable!
Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
to prove such results inside of the guarded setting. 
In particular, if we tried to prove a result such as the above in the guarded
setting, we would run into a problem where we would have a natural number
``stuck'' under a $\later$, with no way to get out the underlying number.

Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
to the more familiar, set-theoretic world with no internal notion of step-indexing.
As mentioned in the Technical Background section (Section \ref{sec:sgdt}), we can do this
using \emph{clock quantification}.

Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
For example, consider the definition of the lift monad by guarded recursion in Section \ref{TODO}.
% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.

Similarly, we can define a ``global'' version of the lock-step error ordering and the
weak bisimilarity relation on elements of the global lift; the former is defined by

\[ x \le^{gl} y := \forall k. x[k] \le y[k], \]

and the latter is defined by

\[ x \bisim^{gl} y := \forall k. x[k] \bisim y[k]. \]

It can be shown that there is an isomorphism between the global lift monad and the
so-called Delay monad of Capretta \cite{lmcs:2265}.
Recall that the delay monad $\text{Delay}(X)$ is defined as the coinductive solution
to the equation

\[ \text{Delay}(X) \cong X + \text{Delay}(X). \]

It can be shown that for a clock-irrelevant type $X$, $\li^{gl} X$ is a final
coalgebra of the functor $F(Y) = X + 1 + Y$ (For example, this follows from Theorem 4.3 in
\cite{kristensen-mogelberg-vezzosi2022}.) Since $\text{Delay}(X + 1)$ is also a final coalgebra
of this functor, then we have $\li^{gl} X \cong \text{Delay}(X + 1)$.

We can define a ``lock-step error ordering'' on $\text{Delay}(X + 1)$

And we similarly define a ``weak bisimilarity'' on $\text{Delay}(X + 1)$

We claim that modulo the above isomorphism, the ``global" version of the
extensional error ordering on $\li^{gl} X$ is equivalent to

% We have been writing the type as $\li X$, but it is perhaps more accurate to write it as $\li^k X$ to
% emphasize that the construction is parameterized by a clock $k$.

>>>>

  We define $\push_{Fc} : \mathbb{N} \times P_{A} \to \mathbb{N} \times P_{A'}$ by

  \[ \push_{Fc}(n, a) = (n, \push_c(a)). \]

  Then we need to build the following square:

  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGQSJdLFsxLDAsIkZBJyJdLFswLDEsIkZBIl0sWzEsMSwiRkEnIl0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCIoXFxkZWx0YV97RkF9XiopXm4gXFxjaXJjIFxccHRiX0EoYSkiLDJdLFsxLDMsIihcXGRlbHRhX3tGQSd9XiopXm4gXFxjaXJjIFxccHRiX3tBJ30oXFxwdXNoX2MoYSkpIl1d
  \[\begin{tikzcd}[ampersand replacement=\&]
    FA \& {FA'} \\
    FA \& {FA'}
    \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \arrow["Fc", "\shortmid"{marking}, no head, from=1-1, to=1-2]
    \arrow["{(\delta_{FA}^*)^n \circ \ptb_A(a)}"', from=1-1, to=2-1]
    \arrow["{(\delta_{FA'}^*)^n \circ \ptb_{A'}(\push_c(a))}", from=1-2, to=2-2]
  \end{tikzcd}\]

  This is obtained by composing the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
  with the square corresponding to the push property for $B$

  % Given $n \in \mathbb{N}$ we define the push to be $n$.
  % Then the required square is $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$,
  % which exists by the definition of a step-1 model.
  % On the other hand, given $a \in P_{A}$, we define the push to be $\push_c(a)$
  % and then the required exists by the push property for $c$.
  % Defining $\pull_{Fc}$ is symmetric.

@@

  % Likewise, given push-pull structures 
  % $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
  % $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
  % we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
  % with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
\end{lemma}
\begin{proof}
We define the push function for $\piv_{c_1 \times c_2}$ by 
$\push(a_1, a_2) = (\push_{c_1}(a_1), \push_{c_2}(a_2))$ and likewise for $\pull$.
The push property holds because it holds for each component.
The pull property is established similarly.
\end{proof}

@@

      \[ (A_1, P_{A_1}, \ptb_{A_1}) \times (A_2, P_{A_2}, \ptb_{A_2}) =
        (A_1 \times A_2, P_{A_1} \times P_{A_2}, \ptb_{A_1 \times A_2}) \]

      where $\ptb_{A_1 \times A_2}(p_1, p_2) = \ptb_{A_1}(p_1) \times \ptb_{A_2}(p_2)$.

@@

      \[ (A, P_A, \ptb_A) \arr (B, P_B, \ptb_B) = 
        (A \arr B, P_A^{op} \times P_B, \ptb_{A \arr B}), \]

@@

    We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
    to a perturbation in $P_{U(A \to B)}$, as follows:
    Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
    $(n, \id_{P_A}, b) \in P_{U(A \to B)}$.
    We need to show that
%
    \[ \ptb_{U(A \arr B)}(\id \tok (n, b)) = \id \tok \ptb_{UB}(n, b). \]
%
    That is, we need to show that
%
    \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \ptb_B(b)) = 
        \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
%
    This follows from the fact that $\delta_{UB}^*$ commutes with computation morphisms.
%

    Given a perturbation $q = (n, a) \in P_{FA}$ we define
    $(n, a) \tok \id = (n, a, \id_{P_B}) \in P_{U(A \to B)}$.
    We need to show that
  %
    \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
  %
    By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes with all computation morphisms.

    We can similarly construct a Kleisli product operation on perturbations.
    Given a perturbation $(n, a_1, a_2) \in P_{F(A_1 \times A_2)}$ we define
    $(n, a_1, a_2) \timesk \id = (n, a_1, \id_{P_{A_2}})$,
    and likewise we define $\id \timesk (n, a_1, a_2) = (n, \id_{P_{A_1}}, a_2)$.

@@

Recall the notion of quasi-order-equivalence as defined in Definition \ref{def:quasi-order-equivalent}.
The following lemma will be useful in showing that two relations are quasi-order-equivalent.

\begin{lemma}\label{lem:left-rep-by-same-morphism}
  Let $c, c' : A \rel A'$. If $c$ and $c'$ are both quasi-left-representable
  by the same $f$, then $c \qordeq c'$.

  Dually, if $d$ and $d'$ are both quasi-right-representable by the same $\phi$, then
  $d \qordeq d'$.
\end{lemma}
\begin{proof}
  By $\upr$ for $c'$, there exists a perturbation $\delle_{c'}$ and a square
  $\upr_{c'} : \delle_{c'} \ltdyn_{c'}^{r(A)} f$.

@@

%%%%%%%%%%%%%%%
% Composition %
%%%%%%%%%%%%%%%
\begin{lemma}\label{lem:representation-comp}
  In the below, let $c : A \rel A'$ and $c' : A' \rel A''$ and $d : B \rel B'$ and $d' : B' \rel B''$.
  \begin{enumerate}
    \item Given left-representation structures $\rho^L_c$ for $c$ and $\rho^L_{c'}$ for $c'$,
    we can define a left-representation structure for the composition $c \comp c'$
    \item Given right-representation structures $\rho^R_d$ for $d$ and $\rho^R_{d'}$ for $d'$,
    we can define a right-representation structure for the composition $d \comp d'$
    \item Given right-representation structures $\rho^R_{Fc}$ for $Fc$ and $\rho^R_{Fc'}$ for $Fc'$,
    we can define a right-representation structure $\rho^R_{F(c \comp c')}$ for $F(c \comp c')$.
    \item Given left-representation structures $\rho^L_{Ud}$ for $Ud$ and $\rho^L_{Ud'}$ for $Ud'$,
    we can define a left-representation structure $\rho^L_{U(d \comp d')}$ for $U(d \comp d')$.
  \end{enumerate}

@@

    First, we claim that $F(c \comp c')$ and $Fc \comp Fc'$ are
    both quasi-left-represented by $Fe_{c'} \circ Fe_c$.
    Indeed, we have by part (1) that $e_c' \circ e_c$ left-represents $c \comp c'$,
    and then since $F$ preserves perturbations and squares it follows that
    $F(e_c' \circ e_c) = Fe_{c'} \circ Fe_c$ quasi-left-represents $F(c \comp c')$.
    On the other hand, we also know that $Fe_c$ quasi-left-represents $Fc$ and $Fe_{c'}$
    left-represents $Fc'$, so their composition quasi-left-represents $Fc \comp Fc'$.
    Thus, by Lemma \ref{lem:left-rep-by-same-morphism}, there is a square $\alpha$ of the form

@@

  % TODO check this
  We define $\rho^R_{F(c_1 \times c_2)}$ as follows:
  \begin{itemize}
    \item $p_{F(c_1 \times c_2)} = (p_{Fc_1} \timesk A_2) \circ (A_1' \timesk p_{Fc_2})$
    \item $\dellp_{F(c_1 \times c_2)} = (\dellp_{Fc_1} \timesk A_2) \circ (A_1 \timesk \dellp_{Fc_2})$
    \item $\delrp_{F(c_1 \times c_2)} = (\delrp_{Fc_1} \timesk A_2') \circ (A_1' \timesk \delrp_{Fc_2})$
    \item The commuting squares are obtained via the functorial action of $\timesk$ on the squares for $Fc_1$ and $Fc_2$.
  \end{itemize}
\end{proof}

@@

\subsection{Constructing an Extensional Model}\label{sec:extensional-construction-appendix}

@@

  Recall the extensional model construction outlined in Section \ref{sec:extensional-model-definition}.
  We first establish the representability properties for this model.
  We show the left-representability squares; the right-representability squares are dual.

@@

We observe that the functoriality of the CBPV connectives on relations up to
order-equivalence is a direct consequence of the functoriality of the CBPV connectives
on relations up to quasi-order-equivalence in the step-4 intensional model, since the
perturbations on both sides of the square are by definition bisimilar to the identity.

Likewise, the retraction properties for the relations
$\iarr{}: U(D \to F D) \rel D$, $\inat : \nat \rel D$, and $\itimes : D \times D \rel D$
hold because they held up to delays in the intensional model, and the delays
disappear in the extensional model construction.

@@
